Scalable Extreme Deconvolution

### 2019-09-09 ml4ps_paper/paper.tex
Update results.
Add Appendix B.

Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
% TODO -- is data standardized?
Training was done for 20 epochs.
The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

\begin{table}{}
  \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.  Average over 5 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Method     & K &  Validation     & Test & Time (minutes) \\
    \midrule
    Existing EM                                     & 64  & $-26.12 \pm 0.03$ & - & $60.0 \pm 7.9$\\
    (\citet{bovyExtremeDeconvolutionInferring2011}) & 128 & $-25.95 \pm 0.04$  & - & $101.1 \pm 2.19$    \\
                                                    & 256 & $-25.76 \pm 0.02$ & - & $195.3 \pm 4.08$ \\
    \midrule
    Minibatch EM  & 64  & $-26.04 \pm 0.01$   & - & $26.8 \pm 1.71$ \\
                  & 128 & $-25.92 \pm 0.01$   & - & $44.5 \pm 1.0$      \\
                  & 256 & $-25.83 \pm 0.003$  & - & $51.6 \pm 3.0$ \\
    \midrule
    SGD & 64    & $-26.10 \pm 0.02$  & - &$23.5 \pm 2.34$ \\
        & 128   & $-25.94 \pm 0.01$  & - & $34.0 \pm 0.83$ \\
        & 256   & $-25.84 \pm 0.01$  & - & $36.6 \pm 1.3$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

The question remains of which method should be used for any particular problem?
The existing EM method had comparable log-likelihoods for smaller sizes of $K$, and did marginally better for the largest size, but at the cost of a much longer training time.
The SGD method is faster than the minibatch EM method, especially as $K$ increases, at a cost of having a slightly worse log-likelihood.
The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
In practice we would suggest using the minibatch EM method for relatively smaller datasets, and the SGD method on larger datasets.

@@

\section{Experiment Details}

>>>>

Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
% TODO -- is data standardized?
Training was done for 20 epochs.
The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

\begin{table}{}
  \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.  Average over 10 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lcccc}
      \toprule
      Method     & K &  Validation     & Test & Time (minutes) \\
      \midrule
      Existing EM & 64 & $-26.10 \pm 0.03$ & - & $55.53 \pm 7.17$ \\
      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - & $100.30 \pm 1.96$ \\
       & 256 & $-25.76 \pm 0.02$ & - & $195.45 \pm 3.93$ \\
       & 512 & $-25.67 \pm 0.01$ & - & $385.60 \pm 8.24$ \\
      \midrule
      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - & $30.98 \pm 0.51$ \\
       & 128 & $-25.91 \pm 0.01$ & - & $47.01 \pm 0.67$ \\
       & 256 & $-25.83 \pm 0.00$ & - & $55.18 \pm 2.89$ \\
       & 512 & $-25.80 \pm 0.00$ & - & $100.47 \pm 3.89$ \\
      \midrule
      SGD & 64 & $-25.89 \pm 0.02$ & - & $28.02 \pm 0.77$ \\
       & 128 & $-25.77 \pm 0.02$ & - & $36.44 \pm 0.44$ \\
       & 256 & $-25.67 \pm 0.02$ & - & $40.54 \pm 1.48$ \\
       & 512 & $-25.59 \pm 0.02$ & - & $64.77 \pm 2.91$ \\
      \bottomrule
  \end{tabular}
\end{table}

@@

The question remains of which method should be used for any particular problem?
The SGD method provides marginally better log-likelihood values, is faster to train, and scales better with component size $K$.
In addition, we found it to be more numerically stable during training, making it the obvious choice.

@@

\section{Experiment Details}
\label{apx:repro}

Here we provide specific details of our experiments for reproducibility.

\subsection{Dataset}

From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $2,000,000$.
Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.

\subsection{Initialisation}
For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
Covariances $V_j$ were set to the identity matrix.

\subsection{Training}

All methods were trained for a total of twenty epochs.
Both minibatch methods used a batch size of $500$.
For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.

For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$, following the scheme of the original publication using $W = wI$, where $w = 10^{-3}$ and $I$ is the identity matrix.
